
\section{Category of categories?}

\NotaInterna{Clearly, this section needs more work\dots{}}

Functors can be composed --- and I think at this point it is not a
secret. Take \(\cat C\), \(\cat D\) and \(\cat E\) categories and
functors
\[\cat C \functo F \cat D \functo G \cat E .\]
The sensible way to define the composite functor
\(GF : \cat C \to \cat E\) is mapping the objects \(x\) of \(\cat C\) to
the objects \(GF(x)\) of \(\cat E\), and the morphisms \(f : x \to y\)
of \(\cat C\) to the morphisms \(GF(f) : GF(x) \to GF(y)\) of
\(\cat E\). That being set, the composition is associative and there
is an identity functor too.

There are all the conditions, so what prevents us to consider a
category --- we can call \(\bf Cat\) --- that has categories as objects
and functors as morphisms?

If we work upon NBG, we can think of any proper class as a category,
for this statement have a closer look at
Example~\ref{example:CollectionsAreCats}. What happens now is that the
class of objects of \(\bf Cat\) has an element that is a proper class,
which isn't legal in NBG.

Is a category of {\em locally small} categories and functors
problematic? Take \(\cat C\) such that \(\cat C(a, b)\) is a proper
class for some \(a\) and \(b\) objects: consider \(\cat C{/}{b}\). In
this case \(\obj{\cat C/b}\) is a proper class too, and here we go
again.

Now what? If we stick to NBG, this is a limit we have to take into
account. From now on, \(\bf Cat\) is the category of {\em small}
categories and functors between small categories.


%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "../CT"
%%% TeX-engine: luatex
%%% End:
